Cross-validation of regression models: mixed model examples (edited v2)

### 2023-10-13 scratch/gm_addition_to_cv-edited_v2.Rmd
further code edits for mixed-models vignette section

---
title: "Cross-validation of regression models: mixed model examples (edited)"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validation of regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

# Parameters:
set.seed(9693) # set.seed(123) 
Nb <- 100     # number of groups
Nw <- 5      # number of individuals within groups
Bb <- 0      # between-group regression coefficient on group mean
SDre <- 2.0  # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1  # within group effect of x

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = 20 + seq(1, Nb*Nw)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- 10 +
        Bb * xm +                  # contextual effect
        Bw * (x - xm) +            # within-group effect
        SDre * rnorm(Nb)[group] +  # random level by group
        SDwithin * rnorm(Nb*Nw)    # random error within groups
    }
  )
```

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set with 10 clusters"
plot <- xyplot(y ~ x, data=Data[ 1:50,], group=group,
               ylim=c(0, 20),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot
```

@@

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept only model:
fit0 <- lmer(y ~ 1 + (1|group), Data)

# effect of x and random intercept:
fit1 <- lmer(y ~ x + (1|group), Data)

# effect of x, mean of x, and random intercept:
fit2 <- lmer(y ~ x + xm + (1|group), Data)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
compareCoefs(fit0,  fit1, fit2)
```

@@

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(fit0, re.form = ~ 0)
  fit_mod0.re <- predict(fit0)
  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
  fit_mod1.re <- predict(fit1)
  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
  fit_mod2.re <- predict(fit2)
})

Data_long <- reshape(Data[1:50, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data.10)))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))

Data_long$model <- factor(c("~ 1", "~ 1 + x", "~ 1 + x + xm")
                   [match(Data_long$modelcode, c("mod0","mod1","mod2"))])
```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values with random intercept model"
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
                         effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
                         effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  key=list(
    space="right",
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values with random intercept model and x as a predictor"
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
                         effect == "fe"), 
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
                         effect == "re"), 
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  ylim=c(-20, 40),
  key=list(
    space="right",
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values with random intercept model and x and the group mean of x as predictors"
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" &

                           effect == "fe"), 
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & 
                         effect == "re"), 
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  ylim = c(0, 20),
  key=list(
    space="right",
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

@@

```{r cross-validation}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
modlist <- models("~ 1"=fit0, "~ x"=fit1, "~ x + xm"=fit2)
cvs_cases <- cv(modlist, data=Data, seed=9693)
cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)

cvs_cases <- cvs_cases |> lapply(rbind) |> 
  do.call(rbind, args = _) |> as.data.frame()
cvs_clusters <- cvs_clusters |> lapply(rbind) |> 
  do.call(rbind, args = _) |> as.data.frame()

cvs_cases$modlist <- names(modlist)
cvs_clusters$modlist <- names(modlist)
cvs_cases$cv_method <- "cases"
cvs_clusters$cv_method <- "clusters"

dout <- within(
  as.data.frame(rbind(cvs_cases, cvs_clusters)),
  {
    model <- factor(modlist)
    CV <- as.numeric(as.character(`CV crit`))
  }
)

dout <- dout[order(dout$model), ]

xyplot(CV ~ model, data=dout, groups=cv_method,
       par.settings =
         list(superpose.line=list(lty=c(2, 3), lwd=3), 
              superpose.symbol=list(pch=15:16, 
                                    cex=1.5)),
       type="b",
       auto.key=TRUE,
       scales=list(y=list(log=TRUE)))
```

>>>>

---
title: "Cross-validation of regression models: mixed model examples (edited v2)"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validation of regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

# Parameters:
set.seed(9693) 
Nb <- 100    # number of groups
Nw <- 5      # number of individuals within groups
Bb <- 0      # between-group regression coefficient on group mean
SDre <- 2.0  # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1     # within group effect of x
Ay <- 10    # intercept for response
Ax <- 20    # starting level of x
Nx <- Nw*10 # number of distinct x values

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- Ay +
        Bb * xm +                  # contextual effect
        Bw * (x - xm) +            # within-group effect
        SDre * rnorm(Nb)[group] +  # random level by group
        SDwithin * rnorm(Nb*Nw)    # random error within groups
    }
  )
```

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set, showing first 10 of 100 clusters."
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot
```

@@

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept only:
fit0 <- lmer(y ~ 1 + (1 | group), Data)

# effect of x and random intercept:
fit1 <- lmer(y ~ x + (1 | group), Data)

# effect of x, mean of x, and random intercept:
fit2 <- lmer(y ~ x + xm + (1 | group), Data)
        # equivalent to y ~ I(x - xm) + xm + (1 | group)

# model generating the data (where Bb = 0)
fit3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
compareCoefs(fit0,  fit1, fit2, fit3)
```

@@

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(fit0, re.form = ~ 0)
  fit_mod0.re <- predict(fit0)
  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
  fit_mod1.re <- predict(fit1)
  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
  fit_mod2.re <- predict(fit2)
  fit_mod3.fe <- predict(fit3, re.form = ~ 0)
  fit_mod3.re <- predict(fit3)
})

Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))

Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from model with random intercept and x as a predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from model with random intercept, x, and the group mean of x as predictors."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

```{r plot-fits-mod3}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values from the estimated model generating the data."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

@@

```{r cross-validation-clusters}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with cluster-based resampling."
modlist <- models("~ 1"=fit0, "~ 1 + x"=fit1, 
                  "~ 1 + x + xm"=fit2, "~ 1 + I(x - xm)"=fit3)
cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
```
```{r cross-validation-cases}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case-based resampling."
cvs_cases <- cv(modlist, data=Data, seed=9693)
plot(cvs_cases, main="Model Comparison, Case-Based CV")
```
